\textbf{Differences in COVID-19 vaccination in the province of Ontario across Health Regions and socio-economic strata}

### 2023-05-08 manuscript_v5/04-discussion.qmd
new directory for version, tidy code

% Options for packages loaded elsewhere
\PassOptionsToPackage{unicode}{hyperref}
\PassOptionsToPackage{hyphens}{url}
\PassOptionsToPackage{dvipsnames,svgnames,x11names}{xcolor}
%
\documentclass[
  letterpaper,
  DIV=11,
  numbers=noendperiod]{scrartcl}

\usepackage{amsmath,amssymb}
\usepackage{lmodern}
\usepackage{iftex}
\ifPDFTeX
  \usepackage[T1]{fontenc}
  \usepackage[utf8]{inputenc}
  \usepackage{textcomp} % provide euro and other symbols
\else % if luatex or xetex
  \usepackage{unicode-math}
  \defaultfontfeatures{Scale=MatchLowercase}
  \defaultfontfeatures[\rmfamily]{Ligatures=TeX,Scale=1}
\fi
% Use upquote if available, for straight quotes in verbatim environments
\IfFileExists{upquote.sty}{\usepackage{upquote}}{}
\IfFileExists{microtype.sty}{% use microtype if available
  \usepackage[]{microtype}
  \UseMicrotypeSet[protrusion]{basicmath} % disable protrusion for tt fonts
}{}
\makeatletter
\@ifundefined{KOMAClassName}{% if non-KOMA class
  \IfFileExists{parskip.sty}{%
    \usepackage{parskip}
  }{% else
    \setlength{\parindent}{0pt}
    \setlength{\parskip}{6pt plus 2pt minus 1pt}}
}{% if KOMA class
  \KOMAoptions{parskip=half}}
\makeatother
\usepackage{xcolor}
\setlength{\emergencystretch}{3em} % prevent overfull lines
\setcounter{secnumdepth}{-\maxdimen} % remove section numbering
% Make \paragraph and \subparagraph free-standing
\ifx\paragraph\undefined\else
  \let\oldparagraph\paragraph
  \renewcommand{\paragraph}[1]{\oldparagraph{#1}\mbox{}}
\fi
\ifx\subparagraph\undefined\else
  \let\oldsubparagraph\subparagraph
  \renewcommand{\subparagraph}[1]{\oldsubparagraph{#1}\mbox{}}
\fi

\providecommand{\tightlist}{%
  \setlength{\itemsep}{0pt}\setlength{\parskip}{0pt}}\usepackage{longtable,booktabs,array}
\usepackage{calc} % for calculating minipage widths
% Correct order of tables after \paragraph or \subparagraph
\usepackage{etoolbox}
\makeatletter
\patchcmd\longtable{\par}{\if@noskipsec\mbox{}\fi\par}{}{}
\makeatother
% Allow footnotes in longtable head/foot
\IfFileExists{footnotehyper.sty}{\usepackage{footnotehyper}}{\usepackage{footnote}}
\makesavenoteenv{longtable}
\usepackage{graphicx}
\makeatletter
\def\maxwidth{\ifdim\Gin@nat@width>\linewidth\linewidth\else\Gin@nat@width\fi}
\def\maxheight{\ifdim\Gin@nat@height>\textheight\textheight\else\Gin@nat@height\fi}
\makeatother
% Scale images if necessary, so that they will not overflow the page
% margins by default, and it is still possible to overwrite the defaults
% using explicit options in \includegraphics[width, height, ...]{}
\setkeys{Gin}{width=\maxwidth,height=\maxheight,keepaspectratio}
% Set default figure placement to htbp
\makeatletter
\def\fps@figure{htbp}
\makeatother
\newlength{\cslhangindent}
\setlength{\cslhangindent}{1.5em}
\newlength{\csllabelwidth}
\setlength{\csllabelwidth}{3em}
\newlength{\cslentryspacingunit} % times entry-spacing
\setlength{\cslentryspacingunit}{\parskip}
\newenvironment{CSLReferences}[2] % #1 hanging-ident, #2 entry spacing
 {% don't indent paragraphs
  \setlength{\parindent}{0pt}
  % turn on hanging indent if param 1 is 1
  \ifodd #1
  \let\oldpar\par
  \def\par{\hangindent=\cslhangindent\oldpar}
  \fi
  % set entry spacing
  \setlength{\parskip}{#2\cslentryspacingunit}
 }%
 {}
\usepackage{calc}
\newcommand{\CSLBlock}[1]{#1\hfill\break}
\newcommand{\CSLLeftMargin}[1]{\parbox[t]{\csllabelwidth}{#1}}
\newcommand{\CSLRightInline}[1]{\parbox[t]{\linewidth - \csllabelwidth}{#1}\break}
\newcommand{\CSLIndent}[1]{\hspace{\cslhangindent}#1}

%line numbers
%\usepackage{mathpazo}
\usepackage{lineno}
\linenumbers

\usepackage{booktabs}
\usepackage{longtable}
\usepackage{array}
\usepackage{multirow}
\usepackage{wrapfig}
\usepackage{float}
\usepackage{colortbl}
\usepackage{pdflscape}
\usepackage{tabu}
\usepackage{threeparttable}
\usepackage{threeparttablex}
\usepackage[normalem]{ulem}
\usepackage{makecell}
\usepackage{xcolor}
\usepackage{amsmath}
\usepackage{caption}
\KOMAoption{captions}{tableheading}
\makeatletter
\makeatother
\makeatletter
\makeatother
\makeatletter
\@ifpackageloaded{caption}{}{\usepackage{caption}}
\AtBeginDocument{%
\ifdefined\contentsname
  \renewcommand*\contentsname{Table of contents}
\else
  \newcommand\contentsname{Table of contents}
\fi
\ifdefined\listfigurename
  \renewcommand*\listfigurename{List of Figures}
\else
  \newcommand\listfigurename{List of Figures}
\fi
\ifdefined\listtablename
  \renewcommand*\listtablename{List of Tables}
\else
  \newcommand\listtablename{List of Tables}
\fi
\ifdefined\figurename
  \renewcommand*\figurename{Figure}
\else
  \newcommand\figurename{Figure}
\fi
\ifdefined\tablename
  \renewcommand*\tablename{Table}
\else
  \newcommand\tablename{Table}
\fi
}
\@ifpackageloaded{float}{}{\usepackage{float}}
\floatstyle{ruled}
\@ifundefined{c@chapter}{\newfloat{codelisting}{h}{lop}}{\newfloat{codelisting}{h}{lop}[chapter]}
\floatname{codelisting}{Listing}
\newcommand*\listoflistings{\listof{codelisting}{List of Listings}}
\makeatother
\makeatletter
\@ifpackageloaded{caption}{}{\usepackage{caption}}
\@ifpackageloaded{subcaption}{}{\usepackage{subcaption}}
\makeatother
\makeatletter
\@ifpackageloaded{tcolorbox}{}{\usepackage[many]{tcolorbox}}
\makeatother
\makeatletter
\@ifundefined{shadecolor}{\definecolor{shadecolor}{rgb}{.97, .97, .97}}
\makeatother
\makeatletter
\makeatother
\ifLuaTeX
  \usepackage{selnolig}  % disable illegal ligatures
\fi
\IfFileExists{bookmark.sty}{\usepackage{bookmark}}{\usepackage{hyperref}}
\IfFileExists{xurl.sty}{\usepackage{xurl}}{} % add URL line breaks if available
\urlstyle{same} % disable monospaced font for URLs
\hypersetup{
  pdftitle={Differences in COVID-19 vaccination in the province of Ontario across Health Regions and socio-economic strata},
  pdfauthor={Ariel Mundo Ortiz1,2; Bouchra Nasri1,2,},
  colorlinks=true,
  linkcolor={blue},
  filecolor={Maroon},
  citecolor={Blue},
  urlcolor={Blue},
  pdfcreator={LaTeX via pandoc}}

\title{\textbf{Differences in COVID-19 vaccination in the province of
Ontario across Health Regions and socio-economic strata}}
\author{Ariel Mundo Ortiz\textsuperscript{1,2} \and Bouchra
Nasri\textsuperscript{1,2,*}}
\date{}

\begin{document}
\maketitle
\ifdefined\Shaded\renewenvironment{Shaded}{\begin{tcolorbox}[borderline west={3pt}{0pt}{shadecolor}, breakable, frame hidden, interior hidden, enhanced, boxrule=0pt, sharp corners]}{\end{tcolorbox}}\fi

\textsuperscript{1} Centre de Recherches Mathématiques, University of
Montreal, Montréal, Canada\\
\textsuperscript{2} Department of Social and Preventive Medicine, École
de Santé Publique, University of Montreal, Montréal, Canada

\textsuperscript{*} Correspondence:
\href{mailto:bouchra.nasri@umontreal.ca}{Bouchra Nasri
\textless{}bouchra.nasri@umontreal.ca\textgreater{}}

\hypertarget{abstract}{%
\section{Abstract}\label{abstract}}

The COVID-19 pandemic continues to be a worldwide public health concern.
Although vaccines against this disease were rapidly developed,
vaccination uptake has not ben equal across all the segments of the
population. In particular, it has been shown that there have been
differences in vaccine uptake across different segments of the
population. However, there are also differences in vaccination across
geographical areas, which might be important to consider in the
development of future public health policies against COVID-19. In this
study, we examined the relationship between vaccination status (having
received the first dose of a COVID-19 vaccine), and different
socio-economic and geographical factors. Our results show that during
the last three months of 2021, individuals in certain equity-deserving
groups (visible minorities) were three times less likely to be
vaccinated than White/Caucasian individuals across the province and that
in some cases, within these groups individuals in low income brackets
had significantly higher odds of vaccination when compared to their
peers in high income brackets. Finally, we identified significantly
lower odds of vaccination in the West Health Region of Ontario within
certain equity-deserving groups. This study shows that there is an
ongoing need to better understand and address differences in vaccination
uptake across diverse segments of the population of Ontario that have
been largely impacted by the pandemic.

\hypertarget{keywords}{%
\section*{Keywords}\label{keywords}}
\addcontentsline{toc}{section}{Keywords}

Covid-19, vaccination, survey, socio-economic factors, visible
minorities.

\hypertarget{background}{%
\section{Background}\label{background}}

The vaccines against COVID-19 have been considered a major achievement
of modern medicine as their rapid development allowed the start of broad
vaccination campaigns towards the end of 2020 in certain countries, such
as the US and
Canada\textsuperscript{\protect\hyperlink{ref-davis2022}{1}--\protect\hyperlink{ref-tanne2020}{3}}.
This made some believe that vaccines were destined to be a determinant
factor in a rapid ending of the
pandemic\textsuperscript{\protect\hyperlink{ref-thelancet2021}{4}}.
However, although it has been estimated that COVID-19 vaccines have
prevented around 14 million of deaths
worldwide\textsuperscript{\protect\hyperlink{ref-watson2022}{5}}, their
implementation has been far from being equal to that of the smallpox and
polio vaccines, which were implemented on a global scale and that were
crucial to control these
diseases\textsuperscript{\protect\hyperlink{ref-kayser2021}{6}}. In
fact, the rollout of COVID-19 vaccines has faced multiple challenges
since its inception which ultimately have hampered their use to achieve
the ultimate goal of global immunity.

This problematic in the rollout of the COVID-19 vaccines is a
multifaceted issue resulting from, among other things, the development
of new variants due to inadequate public health
measures\textsuperscript{\protect\hyperlink{ref-li2021}{7}}, inequality
in vaccine access between high- and low-income
countries\textsuperscript{\protect\hyperlink{ref-gerretsen2021}{8},\protect\hyperlink{ref-tamey2022}{9}},
vaccine
hesitancy\textsuperscript{\protect\hyperlink{ref-nafilyan2021}{10}}, and
differences in vaccination uptake across different segments of the
population\textsuperscript{\protect\hyperlink{ref-malik2020}{11}}. In
particular, it is well established that differences in vaccination
uptake have been present even in countries that have had ample access to
vaccines since 2020 (such as the US, the UK, and Canada), where lower
vaccine uptake has been observed within certain racial groups (i.e.,
individuals that identify as Black, Asian, or Indigenous), and in
individuals within low income
brackets\textsuperscript{\protect\hyperlink{ref-willis2021}{12}--\protect\hyperlink{ref-khubchandani2021}{15}}.
Reasons given for lower vaccine uptake in these cases have included
medical mistrust due to systemic medical
racism\textsuperscript{\protect\hyperlink{ref-stoler2021}{14}}, mistrust
in vaccines\textsuperscript{\protect\hyperlink{ref-willis2021}{12}}, and
the influence of conspiracy
theories\textsuperscript{\protect\hyperlink{ref-bogart2021}{16}--\protect\hyperlink{ref-freeman2020}{18}}.
Moreover, in the case of Canada, lower vaccine uptake has been observed
in young individuals, those with a low educational level, households
with children, those without a regular healthcare provider, individuals
that identify as part of certain equity-deserving groups, and those with
a low household
income\textsuperscript{\protect\hyperlink{ref-guay2022}{19}--\protect\hyperlink{ref-hussain2022}{21}}.

However, it is important to consider that vaccination uptake can also be
influenced by geographical (spatial) factors. In this regard,
differences in COVID-19 vaccination rates have been associated with
varied regional attitudes towards
vaccination\textsuperscript{\protect\hyperlink{ref-malik2020}{11}},
spatial differences in vaccine access and supply, vaccination location
availability, and lack of prioritization of areas where vulnerable
groups
reside\textsuperscript{\protect\hyperlink{ref-bogoch2022}{2},\protect\hyperlink{ref-nguyen2021}{22}}.
Other studies have also shown heterogeneity in vaccine uptake within
small governmental administrative units such as
counties\textsuperscript{\protect\hyperlink{ref-mollalo2021}{23}--\protect\hyperlink{ref-bhuiyan2022}{26}},
and that and that accounting for geographical differences in vaccination
can help predict patterns of booster
uptake\textsuperscript{\protect\hyperlink{ref-wood2022}{27}}. Overall,
the evidence provided by the literature demonstrates the existence of
spatially-driven heterogeneities in vaccine uptake that be used by
decision-makers in the development of public health policies that are
focused on addressing these disparities within specific administrative
or geographical areas.

However, such analyses have been carried mostly in territories outside
of Canada, where available studies have been focused in certain cities
(such as Toronto\textsuperscript{\protect\hyperlink{ref-choi2021}{28}},
or Montreal\textsuperscript{\protect\hyperlink{ref-mckinnon2021}{29}}),
or have explored differences at a province-wide
level\textsuperscript{\protect\hyperlink{ref-guay2022}{19}}. Thus, there
is a need for studies that explore spatial differences in vaccination
within the Canadian territory and that consequently, can help identify
disparities that need to be addressed within specific areas in each
province.

This need is particularly important in the case of Ontario, the most
populated province in Canada. Between 2006 and 2019, Ontario provided
healthcare access to its inhabitants using 14 intra-provincial divisions
called the Local Health Integrated Networks (LHINs). However, this
approach was complex, bureaucratic, and led to systemic
inequalities\textsuperscript{\protect\hyperlink{ref-tsasis2012}{30}}. In
late 2019, the 14 LHINs were phased out and the areas they covered were
incorporated into 6 Health Regions (North East, North West, Central,
Toronto, West, and East) in an effort to improve the healthcare system
of the province\textsuperscript{\protect\hyperlink{ref-dong2022}{31}}.
Because the adoption of the Health Regions is relatively recent, there
is an ongoing need to analyze the impact of this measure and identify
disparities in health access that might exist across the Health Regions,
which can be specially important in the context of the COVID-19
pandemic.

Therefore, in this study we hypothesized that there were differences in
vaccination uptake between the different Health Regions of Ontario
during the last quarter of 2021. By including socio-economic factors in
our analysis, we aimed at identifying in which groups these differences
were significant in order to provide an assessment of the current state
of healthcare access in Ontario.

\hypertarget{methods}{%
\section{Methods}\label{methods}}

\hypertarget{sec-data}{%
\subsection{Data and Methods}\label{sec-data}}

We used data from the \emph{Survey of COVID-19 related Behaviours and
Attitudes}, a repeated cross sectional survey focused on the Canadian
province of Ontario that was commissioned by the Fields Institute for
Research in Mathematical Sciences and the Mathematical Modelling of
COVID-19 Task Force under ethical guidance from the University of
Toronto, and which ran between September 30th, 2021 and January
17th,2022. The survey collected socio-economic information from
participants (Table~\ref{tbl-descriptive-stats}), their location
(nearest municipality, as shown in Figure~\ref{fig-map}), the date of
access to the survey, and asked information on vaccination status by
using the question ``Have you received the first dose of the COVID
vaccine?'', with possible answers ``yes'' and ``no''. The original
dataset contained 39,029 observations (where each observation
corresponded to a unique respondent).

Preliminary analyses of the data included the removal of outliers, of
observations where respondents did not provide answers in all the
covariates of interest, matching the city of each observations with its
corresponding LHIN and Health Region, and removing observations from
areas with low representation (107 observations corresponding to the
North West and North East Health Regions). After all the preliminary
analyses indicated above, the total number of observations used for
analysis was 3,549 which included the East, Central, Toronto, and West
Health Regions covering the period between October 1st,2021 and December
12th, 2021. The original dataset, clean dataset, and details on the data
cleaning process are described in detail in the
\href{https://github.com/aimundo/Fields_COVID-19/}{GitHub repository}
for this paper.

\begin{figure}

{\centering \includegraphics{../data/map_data/map.pdf}

}

\caption{\label{fig-map}Geographic representation of the data collected
by the \emph{Survey of COVID-19 related Behaviours and Attitudes},
collected by the Fields Institute in Ontario. The municipalities
(cities) from where survey participants provided answers (in the clean
dataset) appear as points. The Health six Regions are color-coded.
Internal boundaries within certain Health Regions indicate areas that
belonged to the Local Integrated Health Networks (LHINs), the geographic
areas for healthcare in Ontario before the adoption of the Health
Regions.}

\end{figure}

\hypertarget{statistical-analyses}{%
\subsection{Statistical analyses}\label{statistical-analyses}}

Comparisons between the variables of the survey were performed using the
\(\chi^2\) test. We used a logistic regression model to examine the
impact of the Health Regions in vaccination rates while considering the
socio-economic factors and and months covered by the survey
(Table~\ref{tbl-descriptive-stats}) and certain interactions (Race and
Health Region and Race and income), as previous studies have shown that
socio-economic factors and their interactions are significant predictors
of intent of vaccination and vaccination
status\textsuperscript{\protect\hyperlink{ref-nguyen2022}{32}--\protect\hyperlink{ref-cnat2022a}{34}}.
Because we identified differences in representativity between the survey
data and the estimates from the Census, we used an iterative
proportional fitting procedure
(\emph{raking})\textsuperscript{\protect\hyperlink{ref-deming1940}{35}}
to correct the data using data from the Census and Health Region
population totals; and fitted the regression model to the uncorrected
and corrected data. The reference groups in the model were: 16-34 years
(age group), October(Month), CAD 60,000 and above (income),
White/Caucasian (Race), Toronto (Health Region). Details regarding the
correction can be found in the Appendix. All analyses were conducted in
R 4.2.2 using the packages
\texttt{survey}\textsuperscript{\protect\hyperlink{ref-lumley2011}{36}},\texttt{tidyverse}\textsuperscript{\protect\hyperlink{ref-wickham2019}{37}},
\texttt{quarto}\textsuperscript{\protect\hyperlink{ref-quarto}{38}},
\texttt{modelsummary}\textsuperscript{\protect\hyperlink{ref-modelsummary}{39}},
and
\texttt{gtsummary}\textsuperscript{\protect\hyperlink{ref-gtsummary}{40}}.

\hypertarget{results}{%
\section{Results}\label{results}}

\hypertarget{sample-characteristics}{%
\subsection{Sample Characteristics}\label{sample-characteristics}}

Table~\ref{tbl-descriptive-stats} shows the characteristics of the data
from the Fields COVID-19 survey used for analysis with regard to
vaccination status. A total of 3,549 observations were used for
analysis, from which 27\% (958) corresponded to individuals that
reported not having received the first dose of the vaccine. Respondents
between 16 and 34 years of age, and those that lived within the Health
Region of Toronto had the highest number of observations in the data
(42.8\% and 37.3\% of the total, respectively). Moreover, individuals
who identified as White or Caucasian reported the highest vaccination
uptake (82\%), in contrast to respondents who identified as which had
had the lowest vaccination uptake rate (60\%) across all groups. From
all the covariates, significant associations were identified between
income, race, and vaccination rate.

\hypertarget{tbl-descriptive-stats}{}
\setlength{\LTpost}{0mm}
\begin{longtable}{lccc}
\caption{\label{tbl-descriptive-stats}Descriptive Statistics of the Fields COVID-19 Survey (by Vaccination
Status) }\tabularnewline

\toprule
\textbf{Variable} & \textbf{no}, N = 958\textsuperscript{1} & \textbf{yes}, N = 2,591\textsuperscript{1} & \textbf{p-value}\textsuperscript{2} \\ 
\midrule
Income (CAD) &  &  & <0.001 \\ 
    60000 and above & 305 (23\%) & 1,048 (77\%) &  \\ 
    25000-59999 & 253 (28\%) & 636 (72\%) &  \\ 
    under 25000 & 400 (31\%) & 907 (69\%) &  \\ 
Age Group &  &  & 0.7 \\ 
    16-34 & 409 (27\%) & 1,111 (73\%) &  \\ 
    35-54 & 252 (26\%) & 712 (74\%) &  \\ 
    55 and over & 297 (28\%) & 768 (72\%) &  \\ 
Health Region &  &  & 0.14 \\ 
    Toronto & 371 (28\%) & 952 (72\%) &  \\ 
    Central & 224 (28\%) & 580 (72\%) &  \\ 
    East & 135 (23\%) & 448 (77\%) &  \\ 
    West & 228 (27\%) & 611 (73\%) &  \\ 
Month &  &  & 0.4 \\ 
    October & 469 (27\%) & 1,263 (73\%) &  \\ 
    November & 376 (28\%) & 980 (72\%) &  \\ 
    December & 113 (25\%) & 348 (75\%) &  \\ 
Race &  &  & <0.001 \\ 
    White/Caucasian & 233 (18\%) & 1,079 (82\%) &  \\ 
    Arab/Middle Eastern & 76 (36\%) & 138 (64\%) &  \\ 
    Black & 114 (38\%) & 184 (62\%) &  \\ 
    East Asian/Pacific Islander & 69 (23\%) & 234 (77\%) &  \\ 
    Indigenous & 76 (40\%) & 115 (60\%) &  \\ 
    Latin American & 69 (38\%) & 111 (62\%) &  \\ 
    Mixed & 105 (34\%) & 205 (66\%) &  \\ 
    Other\textsuperscript{3} & 128 (35\%) & 239 (65\%) &  \\ 
    South Asian & 88 (24\%) & 286 (76\%) &  \\ 
\bottomrule
\end{longtable}
\begin{minipage}{\linewidth}
\textsuperscript{1}n (\%)\\
\textsuperscript{2}Pearson\textquotesingle{}s Chi-squared test\\
\textsuperscript{3}Southeast Asian, Filipino, West Asian,
and minorities not identified elsewhere according to the Census.\\
\end{minipage}

\hypertarget{multivariate-regression}{%
\subsection{Multivariate Regression}\label{multivariate-regression}}

Figure~\ref{fig-model-uncorr} shows the estimates from the logistic
regression model of vaccination status for the uncorrected data, whereas
the estimated obtained from the corrected data appear in
Figure~\ref{fig-model-corr}. The results show significantly lower odds
of vaccination in individuals with a low household income and those that
identified as part of equity-deserving groups when compared to
individuals in high income brackets or that identified as White or
Caucasian.

\begin{figure}

\includegraphics{main_files/figure-pdf/fig-model-uncorr-1.pdf} \hfill{}

\caption{\label{fig-model-uncorr}Coefficient estimates and confidence
intervals for the uncorrected model. Only statistically significant
interaction terms are shown. Full interaction terms can be found in
Supplementary Figure A-3.}

\end{figure}

Specifically, significantly lower odds of vaccination were identified
for those with a household income under CAD 25,000 (OR=0.37,
CI={[}0.25,0.56{]}) and those with an income between CAD 25,000 and
59,999 (OR=0.59, CI={[}0.39,0.88{]}). Additionally, individuals who
identified as Arab/Middle Eastern, Black, or Latin American, or that
belonged to ``Other'' racial groups, which included the Southeast Asian,
Filipino, West Asian, and Minorities Not Identified Elsewhere groups
according to the Census, had significantly lower odds of vaccination
than those in the White/Caucasian group (ORs=0.31, 0.32, 0.27, 0.22, and
CIs={[}0.14,0.68{]}, {[}0.17,0.60{]}, {[}0.11,0.66{]}, and
{[}0.12,0.41{]}, respectively). Regarding Health Regions, individuals
that reported living in the West Health Region (which includes the
regions of Waterloo and Niagara, the counties of Wellington, Essex, and
Lambton, and the cities of Hamilton, Haldimand, Brant, and Chatham-Kent)
had significantly higher odds of vaccination than those in the Health
Region of Toronto (OR=1.54, CI={[}1.04,2.29{]}).

Interestingly, individuals in equity-deserving groups with a household
income below CAD 25,000 had higher odds of vaccination (when compared to
those with a household income above CAD 60,000). This held true in the
case of Arab/Middle Eastern (OR=3.08, CI={[}1.27,7.47{]}), Black
(OR=3.15, CI={[}1.43,6.92{]}), and Latin American (OR=2.81,
CI={[}1.04,7.59{]}) individuals, as well as respondents who belonged to
``Other'' minority groups (OR=4.63, CI={[}2.34,9.13{]}), who also had
higher odds of vaccination in the CAD 25,000-59,999 income bracket
(OR=6.96, CI={[}2.67,18.16{]}). Finally, significantly lower odds of
vaccination were identified (when compared to the Toronto Health Region)
for Black individuals in the Central Health Region, which comprises the
region of York, counties of Dufferin and Simcoe and the district of
Muskoka (OR=0.44, CI={[}0.19,0.99{]}), and in individuals that
identified as part of other racial minorities or South Asian that lived
in the West Health Region (ORs=0.41, and CIs={[}0.18,0.92{]} and
{[}0.18,0.95{]}, respectively).

\begin{figure}

\includegraphics{main_files/figure-pdf/fig-model-corr-1.pdf} \hfill{}

\caption{\label{fig-model-corr}Coefficient estimates and confidence
intervals for the corrected model. Only statistically significant
interaction terms are shown. Full interaction terms can be found in
Supplementary Figure A-4.}

\end{figure}

\hypertarget{discussion}{%
\section{Discussion}\label{discussion}}

The existence of healthcare disparities in Ontario motivated the recent
change in the healthcare system of the province, which switched from the
LHIN model to the Health Region model in late
2019\textsuperscript{\protect\hyperlink{ref-tsasis2012}{30},\protect\hyperlink{ref-dong2022}{31}}.
This change is relatively recent, and because of this, is likely that
there are ongoing differences in healthcare access across the province.
In this study we hypothesized that differences in COVID-19 vaccination
uptake were present between the Health Regions, aiming at determining
which socio-demographic groups could be impacted by these disparities
and to provide decision-makers with information that could be used to
develop policies focused on reducing or eliminating these differences
and ensure that the Health Region model is able to improve health access
for the inhabitants of Ontario, a province that faces unique challenges
due to its condition as the most populated and the most ethnically
diverse province of Canada.

Our results indicate that across the most densely populated Health
Regions of Ontario, almost three quarters of surveyed individuals
reported to have received the first dose of the COVID-19 vaccine
(Table~\ref{tbl-descriptive-stats}). We identified significant
intra-provincial differences in vaccination based on socio-economic and
geographical factors. First, our results show differences in odds of
vaccination in individuals with a household income below CAD 60,000 and
in individuals belonging to visible minority groups. Those who
identified as Arab/Middle Eastern, Black, Latin American, or that
belonged to a minority group not included in the survey (Southeast
Asian, Filipino, West Asian, and minority groups not identified
elsewhere) had vaccination odds that were less than a third of
individuals that identified as White/Caucasian
(Figure~\ref{fig-model-corr}). These results are consistent with other
studies that have shown lower vaccination rates in individuals that
identify as part of a racial minority, or that have a low household
income\textsuperscript{\protect\hyperlink{ref-guay2022}{19}--\protect\hyperlink{ref-hussain2022}{21},\protect\hyperlink{ref-carter2022}{41}}.
In the case of Ontario, a possible rationale for this difference is
vaccine access, which has been identified as an important contributing
factor of disparities in vaccination for equity-deserving
groups\textsuperscript{\protect\hyperlink{ref-njoku2021}{42}}, and that
occurred in Toronto earlier in 2021, where it has been shown that
vaccination venues were scarce in low socio-economic areas that had the
highest burden of COVID-19 at the
time\textsuperscript{\protect\hyperlink{ref-bogoch2022}{2}}.

@@

This result is likely due in part to the fact that individuals that
belong racial minority groups tend to perform occupations that have been
deemed as ``essential'' in the context of the
pandemic\textsuperscript{\protect\hyperlink{ref-hawkins2020}{43},\protect\hyperlink{ref-ct2021}{44}},
which include grocery store workers, gas station, warehouse,
distribution, and manufacturing workers, all being occupations for which
an income within the significant brackets identified in the analysis is
to be expected. In the case of Ontario, essential workers had priority
for COVID-19
vaccination\textsuperscript{\protect\hyperlink{ref-mishra2021}{45}},
which would explain the higher odds of vaccination for these
individuals. In other words, it is possible that the type of occupation
from individuals in equity-deserving groups played an important role in
increasing the odds of vaccination.

Additionally, significant higher vaccination odds were identified in the
West Health Region when compared to the Health Region of Toronto
(Figure~\ref{fig-model-corr}). The West Health Region comprises the
regions of Waterloo and Niagara, the counties of Wellington, Essex and
Lambton, and the cities of Hamilton, Haldimand, Brant, and Chatham-Kent.
In this case, these results could reflect the fact that in the survey,
about half of the entries for this Health Region corresponded to
White/Caucasian individuals, who reported a vaccination rate of 83\%
(Supplementary Table A-6). However, the interaction effect of Health
Region and race was also significant in the case of individuals
identifying as South Asian or other minorities not included in the
survey Figure~\ref{fig-model-corr}. In this case, the results of the
interaction term in the model indicate that the odds of vaccination for
those within the South Asian and Other minority groups in the West
Region decreased when compared to the other Health Regions
(Supplementary Figure A-6).

According to Ontario Health, 13.2\% of the population in the West Health
Region identifies as a visible minority, whereas 2.5\% identifies as
Indigenous\textsuperscript{\protect\hyperlink{ref-ontariohealth}{46}}.
Therefore, the estimated lower odds are likely to be explained from a
socio-economic perspective. In fact, 50\% of the answers from this
region in the survey came from the former LHINs of Hamilton Niagara
Haldimand Brant, and Erie St.~Clair, both which are among the regions of
Ontario with the highest proportion of their population (more than 20\%)
in the lowest income
quintile\textsuperscript{\protect\hyperlink{ref-buajitti2018}{47}}
(Supplementary Table A-7). Interestingly, a disproportionate number of
COVID-19 cases and low vaccination rate (under 50\%) have been also
reported in the South Asian community of
Ontario\textsuperscript{\protect\hyperlink{ref-anand2022}{48}}. These
results provide context to the observed differences in vaccination among
the Health Regions of the province, showing which equity-deserving and
socio-demographic groups in certain regions need to be prioritized in
future vaccination strategies. Moreover, these results provide a
rationale for future studies that explore how vaccination uptake varies
across different minority groups within Ontario and other Canadian
provinces.

>>>>

# Discussion

The existence of healthcare disparities in Ontario motivated the recent change in the healthcare system of the province, which switched from the LHIN model to the Health Region model in late 2019  [@tsasis2012;@dong2022]. This change is relatively recent, and because of this, is likely that there are ongoing differences in healthcare access across the province. In this study we hypothesized that differences in COVID-19 vaccination uptake were present between the Health Regions, aiming at determining which socio-demographic groups could be impacted by these disparities and to provide decision-makers with information that could be used to develop policies focused on reducing or eliminating these differences and ensure that the Health Region model is able to improve health access for the inhabitants of Ontario, a province that faces unique challenges due to its condition as the most populated and the most ethnically diverse province of Canada.

Our results indicate that across the most densely populated Health Regions of Ontario, almost three quarters of surveyed individuals reported to have received the first dose of the COVID-19 vaccine (@tbl-descriptive-stats). We identified significant intra-provincial differences in vaccination based on socio-economic and geographical factors. First, our results show differences in odds of vaccination in individuals with a household income below CAD 60,000 and in individuals belonging to visible minority groups. Those who identified as Arab/Middle Eastern, Black, Latin American, or  that belonged to a minority group not included in the survey (Southeast Asian, Filipino, West Asian, and minority groups not identified elsewhere) had vaccination odds that were less than a third of individuals that identified as White/Caucasian (@fig-model-corr). These results are consistent with other studies that have shown lower vaccination rates in individuals that identify as part of a racial minority, or that have a low household income [@guay2022;@muhajarine2021;@carter2022;@hussain2022]. In the case of Ontario, a possible rationale for this difference is vaccine access, which has been identified as an important contributing factor of disparities in vaccination for equity-deserving groups [@njoku2021], and that occurred in Toronto earlier in 2021, where it has been shown that vaccination venues were scarce in low socio-economic areas that had the highest burden of COVID-19 at the time [@bogoch2022].

@@

This result is likely due in part to the fact that individuals that belong racial minority groups tend to perform occupations that have been deemed as "essential" in the context of the pandemic [@hawkins2020;@ct2021], which include grocery store workers, gas station, warehouse, distribution, and manufacturing workers, all being occupations for which an income within the significant brackets identified in the analysis is to be expected.  In the case of Ontario, essential workers had priority for COVID-19 vaccination [@mishra2021], which would explain the higher odds of vaccination for these individuals. In other words, it is possible that the type of occupation from individuals in equity-deserving groups played an important role in increasing the odds of vaccination.

Additionally, significant higher vaccination odds were identified in the West Health Region when compared to the Health Region of Toronto (@fig-model-corr). The West Health Region comprises the regions of Waterloo and Niagara, the counties of Wellington, Essex and Lambton, and the cities of Hamilton, Haldimand, Brant, and Chatham-Kent. In this case, these results could reflect the fact that in the survey, about half of the entries for this Health Region corresponded to White/Caucasian individuals, who reported a vaccination rate of 83\% (Supplementary Table A-6). However, the interaction effect of Health Region and race was also significant in the case of individuals identifying as South Asian or other minorities not included in the survey [@fig-model-corr]. In this case, the results of the interaction term in the model indicate that the odds of vaccination for those within the South Asian and Other minority groups in the West Region decreased when compared to the other Health Regions (Supplementary Figure A-6).

According to Ontario Health, 13.2% of the population in the West Health Region identifies as a visible minority, whereas 2.5% identifies as Indigenous [@ontariohealth]. Therefore, the estimated lower odds are likely to be explained from a socio-economic perspective. In fact, 50% of the answers from this region in the survey came from the former LHINs of Hamilton Niagara Haldimand Brant, and Erie St. Clair, both which are among the regions of Ontario with the highest proportion of their population (more than 20%) in the lowest income quintile [@buajitti2018] (Supplementary Table A-7). Interestingly, a disproportionate number of COVID-19 cases and low vaccination rate (under 50%) have been also reported in the South Asian community of Ontario [@anand2022]. These results provide context to the observed differences in vaccination among the Health Regions of the province, showing which equity-deserving and socio-demographic groups in certain regions need to be prioritized in future vaccination strategies. Moreover, these results provide a rationale for future studies that explore how vaccination uptake varies across different minority groups within Ontario and other Canadian provinces.
